Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-08-17 manuscript/revision2/response_to_reviews/response_to_reviews.qmd

---
format:
  pdf:
    keep-tex: true
    pdfengine: xelatex
    geometry: 
      - "top=1in"
      - "bottom=1in"
      - "left=1in"
      - "right=1in"
    code-annotations: hover
---

::: {.flushright data-latex="1" linestretch="1"}
The University of Texas at Austin\
Department of Integrative Biology\
2415 Speedway\
Austin, TX 78712
:::

::: {.flushright data-latex="1" linestretch="1"}
August 17th, 2024
:::

@@

> ::: {.callout-warning appearance="minimal"}
> Reviewer #1:\
> \
> Suitable Quality?: No\
> Sufficient General Interest?: No\
> Conclusions Justified?: Yes\
> Clearly Written?: No\
> Procedures Described?: Yes\
> \
> Comments on Significance Statement:\
> \
> Conclusions are very specific for the crop/population and traits. General conclusions are lacking.\
> \
> Comments:\
> \
> The paper by MacQueen et al. presents results of a study on two switchgrass populations grown in 8 garden experiments across a rane of latitudes. Traits are green-up date and flowering time. The data show strong GxE, effects of QTL changing in magnitude and sign. Results are partly validated in an independent segregating population.\
> \
> The topic is timely and highly relevant. A better understanding of GxE will be valuable for plant genetics, evolution and breeding. The data are complex and the authors have succeeded in making them accessible to the reader.\
> \
> \
> Relevant literature\
> The introduction and discussion are focused on findings in Arabidopsis and switchgrass. An excellent study from Arabidopsis with high relevance for this paper is\
> Fournier-Level et al. 2016 <https://doi.org/10.1073/pnas.1517456113>\
> \
> Major findings from other crop species have been ignored. To mention just two references:\
> \
> Bustos-Korts et al. 2019 The Plant Journal <https://doi.org/10.1111/tpj.14414>\
> Millet et al. 2016 Plant Physiology <https://doi.org/10.1104/pp.16.00621> and\
> There are many papers on crop modelling (eg work by Mark Cooper and others in maize) and the authors might want to consider linking their results to crop modelling to demonstrate the relevance of the results of this study
> :::

We now add these citations (lines 51-54; lines 62, 125-131). We regret we can't comprehensively reference the impressive crop modeling literature. Our goal was to reference work on natural populations that focused the genetics on phenology and/or local adaptation, and to touch on how these ideas have interacted with crop improvement. This was because our work uses a natural population of switchgrass with potential relevance for breeding efforts in this species.

@@

> ::: {.callout-warning appearance="minimal"}
> Results\
> \
> There is a large overlap of data and findings with reference 32 Lovell et al. 2021 Nature. It has not become clear which additional insights can be gained from this study. One reason might be that objectives of the study have remained rather vague as well as the generic conclusions that can be drawn from the results.
> :::

While we do rely on the same genotypic data and common gardens used in [Lovell *et al.,* 2021](https://doi.org/10.1038/s41586-020-03127-1), and make use of the population structure findings of [Lovell *et al.,* 2021](https://doi.org/10.1038/s41586-020-03127-1) to divide our genotypes into subpopulations for GWAS, the previous work considered fitness proxies and this work considers phenological traits. This work has two additional, key advances over the previous work: it can assign loci genome-wide to multiple kinds of GxE and GxWeather, and has an unbiased statistical method to identify loci with and without rank-changing GxE. We now summarize the findings from [Lovell *et al.,* 2021](https://doi.org/10.1038/s41586-020-03127-1) that we use as a springboard for this work on lines 142-150, and outline the additional advances of this study on lines 162-169, 175-177, and 884-895.

@@

> ::: {.callout-warning appearance="minimal"}
> In the introduction the aim is given as „we test if these populations differ in their phenological adaptation and hence their phenological GxE". Given prior knowledge such as the origin of the populations and earlier results (Lovell et al.) what is expected?
> :::

We now edit the Introduction to more clearly articulate the aims of this manuscript. Our key expectation is that different genetic subpopulations and genomic regions have likely evolved distinct patterns of GxE (lines 59-62; lines 228-232). Thus, we aim to identify the kinds of GxE present for each subpopulation and each trait. (lines 135-137).

@@

> ::: {.callout-warning appearance="minimal"}
> The results for these two populations did not show a general pattern leading to conclusions on the environmental cues. Associations trait/cue changed across populations, traits, gardens, analysis etc. , it was difficult to see the big picture.
> :::

We agree that we do not identify one general pattern or one environmental cue affecting any set of populations or phenological trait. Instead, our key expectation is that different populations and different genetic loci will have evolved distinct patterns of GxE (lines 59-62; lines 228-232; SI Appendix Section S1). Thus, we aim to identify the kinds of GxE present genome-wide and elaborate when these kinds of GxE differ (Figure 2), and demonstrate that individual loci have different types of GxE across environmental regions and subpopulations (Figure 3).

@@

We edit the Results to (i) change our analysis add an algorithm to more rigorously demonstrate this claim given the data that we have, and (ii) more clearly articulate this claim. We now employ a greedy mash algorithm to iteratively add covariance matrices to the *mash* model, only adding matrices that significantly improve the mash model log-likelihood, as we elaborate on further below in our [response to reviewer 2](#fig-greedy).

We do see ways we could ultimately use this for prediction - shrinking effects so that they have more accurate GxE would improve prediction, as in ([Zhu *et al.*, 2023](https://doi.org/10.1016/j.xgen.2023.100297)). However, prediction is beyond the scope of this manuscript; here we are introducing these ideas and using them to do two specific things: (i) find the GxE structures and (ii) look at the genome-wide prevalence of these structures. We feel that using mash to improve predictions is beyond the scope of this paper.

> ::: {.callout-warning appearance="minimal"}
> Results from the SNPs "mash model of Midwest green-up fell on a covariance matrix of average temperature in the 10 days prior to Midwest green-up". These results should be linked to the phenotypic GxE analysis.
> :::

We edit the results to directly compare the GxWeather matrices and the phenotypic correlations in Figure 1B. (lines 251-269; lines 284-287; lines 290-293; lines 297-301).

@@

The GxE analysis in Figure 3 shows the re-estimated effects of these loci on traits as contrasts between individual pairs of gardens. From Figure 3, you can conclude that many loci have effects with antagonistic pleiotropy at the site level.

> ::: {.callout-warning appearance="minimal"}
> Discussion\
> "we must understand the current patterns of trait covariation across environments, .....". What is the understanding from your data? Are there any general patterns or do we need to assess them individually for different genetic material, environments, traits? It looks like the latter so what are the consequences?
> :::

As we state in the SI Appendix (Section S1, paragraph two), we were interested in the types of genetic correlation that could be expressed by different alleles across the genome. We reasoned that different loci could have different patterns of genetic correlation. For example, the effects of an allele of locus $A$ could covary with a weather-based cue if $A$ had a common allele, $a$, that was responsive to this cue. In contrast, the effects of a different locus $B$ might have an effect at only one garden, for example if a pathogen was only present at one garden, and $B$ has a common allele $b$ that is resistant to that pathogen.

As a consequence, we develop an approach to specify multiple environmental cues and compete them to explain patterns of genetic effects (lines 238-244): we use a greedy algorithm to use with mash to select covariance patterns, and use mash to flexibly identify these patterns across populations and environments (lines 244-250). With this approach, we are able to demonstrate that we can associate multiple patterns of GxWeather with specific genomic regions (lines 427-430). We can also assign genetic effects to both GxWeather patterns derived from weather variables and to other, site-based patterns (lines 430-432).

@@

> ::: {.callout-warning appearance="minimal"}
> "this is the first experimental work using QTL mapping and GWAS across ....". I disagree, see references above and others.
> :::

We edit the manuscript to remove this unclear claim. NB: We do cite the first experimental work using both QTL mapping and GWAS ([Brachi *et al.,* 2010](https://doi.org/10.1371/journal.pgen.1000940)), and there are many others; we meant, more specifically, using these approaches to map multiple types of GxE.

> ::: {.callout-warning appearance="minimal"}
> "Gulf and Midwest subpopulations have two distinct photoperiod-related flowering responses....". This has been shown for other species, eg Unterseer et al. 2016 Genome Biology
> :::

We now reference this paper in the introduction (lines 51-56).

@@

> ::: {.callout-warning appearance="minimal"}
> Data/Methods\
> \
> Is one year data enough to make conclusions about the effect of environmental cues? It is well known that GxYear is more pronounced than GxLocation.
> :::

We agree that this would be an interesting topic for future work. Our aim was to demonstrate the value of this approach for mapping GxE, and we believe one year of data sufficient to do this.

> ::: {.callout-warning appearance="minimal"}
> The reader should be able to understand what the hypothesis in the hypothesis-driven models is without looking at the supplement.
> :::

Agreed. We add Table 1 to better explain the covariance matrices we select using the greedy mash algorithm. We include the matrices selected by this algorithm that have \>0.1% weight (lines 273-274) in the mash model as visualizations in Figure 2.

@@

We thank the reviewer for this insight; we have reframed the way we talk about these covariance matrices in the paper (starting with the title, abstract, and significance statement; continuing throughout). Instead of hypothesis-based covariance matrices, we emphasize that the matrices we create are looking at an interaction between genetics and weather - they are GxWeather matrices, and we look for effects with different GxWeather interactions that are based on specific weather cues.

> ::: {.callout-warning appearance="minimal"}
> Statistical issues:\
> \
> - Narrow-sense heritability: A key result is that the h\^2 is low when measured across all trials, but high within trials. However the model used to estimate the global h\^2 is missing a term for replication of lines across accessions. u only accounts for replication of additive effects of lines across gardens, but non-additive genetics can be persistent too. Also, this model doesn’t allow var(e) to vary across locations but it’s likely that it does. If the goal is to show rank-crossing GxE, actually fit a GxE model like was done for the environmental indices to directly show it.\
> - Variance components analysis: The specification of this model is incorrect because Var(u) and Var(ul) are assigned the same distribution. There are many ways of specifying GxE effects in a model like this, and it’s not clear how it is done here. If G is a nxn GRM for n accessions, it can’t be the covariance matrix for both Var(u) (dimension nxn) and Var(ul) (dimension nl x nl). Generally, Var(ul) would be G \otimes Psi where Psi is a lxl covariance matrix among gardens. Sometimes this is constrained to be diagonal (ie no covariance among gardens), and sometimes to be proportional to I (constant variance and no covariance among gardens), but these more restrictive models should be justified.
> :::

We agree that correctly specifying the type of GxE in our models of additive variation across gardens is important for all models in this manuscript. In many cases the complexity of the genetic variance-covariance structures we wish to fit are not be possible to fit using standard approaches. Eight gardens give eight variance components and 8 x 8 covariances - most datasets are not large enough to give solutions for these models. The standard approach is to either do data reduction (e.g. factor analysis) or impose unrealistic constraints (e.g., equal variances, or only positive covariances) in order to solve these models. Our approach and our focus on mash was to allow us to fit less structured variance-covariance structures, as well as more of them.

@@

> ::: {#comment-cov-correlations .callout-warning appearance="minimal"}
> \- Mash: A lot of the results interpretation is based on interpreting the SNP loadings on the specified covariance matrices. This is a secondary use of mash (the primary being the refinement of effect sizes), and while they do interpret these somewhat in the mash paper, these loadings need to be interpreted with caution. If these covariance matrices are similar, mash will somewhat arbitrarily assign weight to any of the matrices because all lead to the same fit to the effect-size data. The mash algorithm doesn’t give the full posterior distribution on the loadings so you can’t check for posterior correlations there. It seems likely to me that the hypothesis-driven and data-driven covariance matrices are somewhat correlated here, and the correlations may differ between green-up and flowering because of the better correlation between the environmental metrics and flowering than green-up. In the mash paper, they used cross-validation with the likelihood in the testing set as the evaluation metric to compare models. It might be safer to try dropping specific covariance matrices and comparing the model performance in a held-out testing set to evaluate the importance of the hypothesis-derived covariance matrices.
> :::

We thank the reviewer for this comment and agree that the selection of covariance matrices to be included in a mash model is a nontrivial question. To clarify our interpretation of the posterior matrix weights provided by mash, we performed an extensive analysis of the performance of mash models when different covariance matrices are included. Specifically, we implemented a model selection approach that uses a greedy algorithm to evaluate the log likelihood of the mash model as additional covariance matrices were included (see Methods lines 901-908, SI Appendix Section S4, and pseudo code below). This is a similar, but more computationally efficient design, than the leave-one-out approach recommended by the reviewer as the runtime of mash increases with the number of covariance matrices included. Additionally, while cross-validation is a powerful approach to model evaluation our limited sample size was prohibitive to the necessary partitioning of individuals included in our analysis.

@@

We now use the narrow-sense heritability as the diagonal for each GxWeather matrix, and clarify this in the Supplementary Methods (SI Appendix, Section S1, first and eighth paragraph - first and last paragraph on page two).

> ::: {.callout-warning appearance="minimal"}
> \- Antagonistic Pleiotropy: Maybe I’m missing the idea of this analysis, but it doesn’t seem to me that this strategy for partitioning loci between antagonistic pleiotropy and differential sensitivity would have equal power to detect each class. It seems that antagonistic pleiotropy should be a subset of differential sensitivity because the latter is defined as simply a change in magnitude. But even if it’s restricted to “different magnitude but same sign”, the practical definition here seems to be that antagonistic pleiotropy is detected when the signs are different and both lfsrs \< threshold (0.05?), while differential sensitivity additionally requires a threshold on the difference in effect sizes (0.4) with no justification for why this size was chosen. This additional criterion will necessarily make the rate of detection different. Even without this, with differential sensitivity one effect size must be much larger than the other in absolute value (which isn’t the case with antagonistic pleiotropy), so differential sensitivity loci will be more likely to pass the lfsr loci than antagonistic pleiotropy loci. Note that previous literature here tried to separate antagonistic pleiotropy from conditional neutrality (ie where in one location the effect was zero), while here the contrast is with differential sensitivity where effects are non-zero in all locations. I’m not aware of previous discussions in the literature of trying to differentiate these two specific classes of GxE variants and not really sure theoretically what the importance would be.
> :::

We believe that the use of the lfsr to detect effects with rank-changing GxE is a key advance in our manuscript, and so we have made changes to the Introduction (lines 175-177), Results (lines 338-343), and Materials & Methods (lines 959-982; lines 996-998; lines 1055-1057) to explain and motivate this statistical change.

@@

Our expectation is that in nature, effects will always differ. Statistically, this difference will not always be detectable. Thus, rather than ask "Are these two effects different?" - as we reasonably expect two effects to be, even if this difference cannot be measured - the local false sign rate answers a more meaningful question: Can we be confident in the sign of this effect?

We think that if evolutionary geneticists are interested in antagonistic pleiotropy and rank-changing GxE, then they should use the lfsr to measure their confidence in the sign of the effect, rather than using statistical tests where the null hypothesis is that the effect is different than zero. However, if we use the lfsr, comparisons between antagonistic pleiotropy and conditional neutrality are no longer sensible, as we are no longer doing a statistical test that can robustly detect conditional neutrality (just as the FDR does not robustly detect antagonistic pleiotropy). This is justified because detection of antagonistic pleiotropy is more important than detection of conditional neutrality for models of local adaptation.

It's true that the threshold for differential sensitivity is arbitrary. We have reworded the manuscript to stress that there is equal power to detect effects of different sign as there are effects of the same sign (lines 996-998; lines 1055-1057). Another class of effects we have in our analysis that differ from classic antagonistic pleiotropy & conditional neutrality comparisons are effects which are not distinguishable by sign or magnitude. It is not accurate that differentially sensitive loci will be more likely to pass the lfsr, because the loci are first selected by significant lfsr, then divided further into effects that can & cannot be distinguished by magnitude. Thus effects that are significant, but have small magnitude in both conditions will be labeled as not being distinguishable - essentially, these effects do not have detectable GxE. Effects without GxE may be even more common and should also be quantified. We propose that, if our intent is to detect sign-changing GxE, then we should use an unbiased statistical test to look at loci with and without sign-changing GxE, the lfsr.

### 2024-08-17 manuscript/revision2/revision2.qmd
revison2: finalize manuscript and cover letter for resubmission

  - name: Phillip A. Fay

@@

  The timing of vegetative and reproductive growth in plants ("phenological timings") depends on genetic effects (G), environmental (e.g., weather) cues, and their interaction. Measuring GxWeather interactions can aid prediction and manipulation of phenological timings. Here, we map GxWeather effects on phenological timings in two highly divergent switchgrass (*Panicum virgatum*) populations using repeated plantings of cloned individuals from these populations at eight sites across the central United States. Some genetic effects covary with  interpretable weather-based cues. For example, in the Gulf population, 65% of genetic effects on the timing of vegetative growth covary with daylength 14 days prior to green-up date, and 33% of genetic effects on the timing of flowering covary with cumulative rainfall in the seven days prior to flowering. However, most variation in genetic effects is site-specific and cannot be attributed to variation in weather variables. We demonstrate that we can identify genetic variation with GxWeather and assign these loci to specific weather-based cues or other patterns. We predict that selective breeding for particular alleles at the GxWeather loci could alter flowering responsiveness in a photoperiod or rainfall-specific way. More broadly, our approach proposes a refined characterization of genotype-by-environment interactions in any species phenotyped in multiple environments.

@@

  We thank the Brackenridge Field laboratory, the Ladybird Johnson Wildflower Center, and the Juenger laboratory for support with plant care and propagation. This material is based upon work supported in part by the Great Lakes Bioenergy Research Center, U.S. Department of Energy, Office of Science, Office of Biological and Environmental Research under Award Numbers DE-SC0018409 and DE-FC02-07ER64494, the US Department of Energy Awards DESC0014156 to T.E.J., DE-SC0017883 to D.B.L, NIH grant R35GM151108 and a Pew Scholarship to A.H., National Science Foundation PGRP Awards IOS0922457 and IOS1444533 to T.E.J, and the Long-term Ecological Research Program (DEB 1832042) at the Kellogg Biological Station.

@@

\firstpage[17]{2}

@@

Flowering time, or the transition from vegetative to reproductive growth, is a common subject of GxE research [@blackman2013interacting; @henry2014transitions; @aagren2017adaptive; @brachi2010linkage; @dittmar2014flowering; @li2018genomic; @romero2017study; @bustos-korts2019], a key output of selection driving adaptation to local environments [@unterseer2016; @andres2012genetic; @wadgymar2017identifying; @blumel2015flowering; @fournier-level2016], and a selection target for crop improvement to adapt crops to local or future environments @jung2009flowering. Changing flowering responsiveness to photoperiod cues has allowed geographic range expansion and increased yields in several cereal species [@unterseer2016; @turner2005pseudo; @faure2012mutation; @hung2012zmcct; @zakhrabekova2012induced; @yang2013oself3] and other crops [@pin2012multifaceted; @weller2019parallel]. Recent statistical advances in studying phenological GxE have involved determining critical environmental indices before the phenological event occurs, such as photothermal time within a critical growth window @li2018genomic. However, most studies of flowering GxE focus on finding a single, best fitting form of genotype-environment covariance, despite the key expectation that different genetic subpopulations, and even different genomic regions, have likely evolved distinct patterns of GxE [@Weine2023.06.21.545998]. Additionally, despite theoretical predictions that local adaptation should involve antagonistic pleiotropy, or alleles with effects with opposing fitness outcomes [@levene1953genetic; @felsenstein1976theoretical; @kawecki2004conceptual; @hedrick1986genetic], previous empirical work has found limited evidence of antagonistic pleiotropy [@des2013genotype; @wadgymar2017identifying; @millet2016]. However, this work has been limited by a known statistical bias that reduced detection of genetic effects that differ in sign [@des2013genotype; @anderson2013genetic; @anderson2011evolutionary]. Thus, despite substantial interest in the frequencies of various forms of GxE, the frequency of sign-changing GxE relative to other forms of GxE remains unknown.

@@

We expected patterns of GxE and GxWeather to vary at the locus, subpopulation, and trait level. Thus, we were interested in modeling the different ways effects might covary across our eight common gardens. We jointly re-estimated the genetic effects of a subset of SNPs across all eight common gardens using *mash* models, which can flexibly capture many modes of covariance across gardens (SI Appendix, Sections S1-S5, Datasets 1-6). To obtain genetic effects, we used subsets of effects from garden- and subpopulation GWAS (SI Appendix, Section S2-S3; Table S1). To obtain covariance structures, we specified both covariance structures introduced in the initial *mash* manuscript, and GxWeather covariance matrices estimated from the covariance of empirical weather patterns at each garden at specific times before the phenological event (Table 1; SI Appendix, Section S1). Then, we used a greedy algorithm to select covariance matrices from each model's set that significantly improved the model likelihood (SI Appendix, Section S4; Table S2). Finally, we fit *mash* models using the selected set of covariance matrices on a subset relatively unlinked (r^2^ \< 0.2), 'strong' genetic effects with low *p*-values (SI Appendix, Section S5).

@@

For green-up date, SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (@fig-covar B). In total, 65% of the posterior weight of strong SNP effects in the *mash* model of Gulf green-up fell on a covariance matrix constructed using the covariance of day length 14 days prior to the date of green-up. The covariance matrix for this weather cue was visually similar to the observed pattern of phenotypic correlation for the timing of vegetative growth in the Gulf subpopulation (@fig-map B; @fig-covar A). *Mash* models of the timing of vegetative growth in the Midwest subpopulation and Both subpopulations did not include this GxWeather covariance. The Midwest had 0.15% weight on a GxWeather covariance matrix of cumulative GDD for the 28d prior to green-up, a GxWeather matrix that was visually similar to the Midwest's observed pattern of phenotypic correlation (@fig-map B). Both subpopulations had non-zero weights on two additional GxWeather covariance types, average temperature one day prior to green-up, and the day length change in seconds in the day prior to green-up (@fig-covar B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations observed in Both subpopulations (@fig-map B; @fig-covar A). The Gulf subpopulation and Both subpopulations had substantially more mass on GxWeather covariance matrices than the Midwest population (@fig-covar C) for the onset of vegetative growth.

For flowering date, distinct GxWeather covariance structures captured covariance in effect sizes for SNPs in the Gulf and Midwest subpopulations. 33% of SNP effects on flowering in the Gulf subpopulation covaried with cumulative rainfall in the seven days prior to flowering (@fig-covar D). 22.6% of SNP effects on flowering in the Midwest subpopulation covaried with day length change in the two days prior to flowering (@fig-covar D), which had negative covariances between Texas and North gardens. Neither of these covariance matrices significantly improved the log-likelihood of the *mash* model of Both subpopulations, and no GxWeather covariance matrices had non-zero mass in models of Both subpopulations (@fig-covar E). In five of the six *mash* models of strong effects, the GxWeather covariance matrices captured a minority of the posterior weights of the strong effects (@fig-covar C,E); in these five models, the majority of this mass was on various canonical covariance matrices. These matrices included simple heterozygosity, with intermediate, positive covariances between all gardens, and single effect matrices with garden-specific effects.

@@

We expected to observe common rank-changing GxE at the level of individual loci as this is a key theoretical prediction of local adaptation [@felsenstein1976theoretical; @levene1953genetic; @kawecki2004conceptual; @hedrick1986genetic]. To determine the frequency of rank-changing GxE, we used the local false sign rate (lfsr), an analogue of the local false discovery rate that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for a sign change between gardens carry an equal statistical burden to those for effects with the same sign. We separated kinds of effects at the level of individual loci into SNP effects that differ in sign between gardens (effects with rank-changing GxE between gardens), SNP effects that differ in magnitude (effects that are large in one garden or region and smaller in others), and SNP effects that are indistinguishable in two gardens (similarly large or small in comparisons between gardens) (@fig-effects A).

@@

As the climate and the natural environment change, it is increasingly critical to understand how patterns of plant-environment interactions will change in response. To do this, we must understand the current patterns of trait covariation across environments, the genetic underpinnings of these patterns, and the cases where this covariation can be altered. Here, we demonstrate that we can associate multiple patterns of GxWeather with specific genomic regions using a switchgrass diversity panel grown at eight common gardens. We assigned genetic effects to both GxWeather patterns with interpretable weather-based cues, and to unmeasured, site-based patterns. We used this approach to study GxWeather for the timings of vegetative and reproductive development in the deeply genetically diverged Gulf and Midwest subpopulations of switchgrass.

Our analysis of the timing of vegetative and reproductive growth revealed many alleles with rank-changing GxE, or sign changes in their effects between the Texas and North gardens (@fig-effects B). The Gulf & Both subpopulations had rank-changing GxE for green-up date, while the Midwest subpopulation had rank-changing GxE for flowering date. As phenological timings are major components of plant fitness, this result supports theoretical models that local adaptation should involve trade-offs at the level of individual loci [@kawecki2004conceptual; @felsenstein1976theoretical; @levene1953genetic; @hedrick1986genetic]. In addition, it advances local adaptation research by using more than two field sites and a wide range of genetic variation to determine that rank-changing GxE for phenological traits is common in small genomic regions [@wadgymar2017identifying; @savolainen2013ecological; @lowry2019qtl].

>>>>

  - name: Philip A. Fay

@@

  The timing of vegetative and reproductive growth in plants ("phenological timings") depend on genetic effects (G), environmental (e.g., weather) cues, and their interaction. Here, we measure phenological timings in two highly divergent switchgrass (*Panicum virgatum*) subpopulations using repeated plantings of cloned individuals at eight sites across the central United States. The timing of vegetative growth for the two subpopulations reversed between their two natural ranges and had strong negative correlations between these regions; in contrast, the timing of flowering was positively correlated between gardens. We expect that these phenotypic correlations consist of polygenic effects on phenology which have distinct patterns of GxE segregating at different mapped loci. Thus, we infer the mixture of ways genetic effects impact phenological timings, such as across common gardens (GxE) or with weather cues (GxWeather). We demonstrate that we can identify genetic variation with GxWeather and assign genetic loci to specific weather-based cues or other patterns. For example, in the Gulf subpopulation, 65% of genetic effects on the timing of vegetative growth covary with daylength 14 days prior to green-up date, and 33% of genetic effects on the timing of flowering covary with cumulative rainfall in the week prior to flowering. However, most variation in genetic effects cannot be attributed to variation in weather variables. Selective breeding for particular alleles at GxWeather loci could alter flowering responsiveness in a photoperiod or rainfall-specific way. More broadly, our approach refines the characterization of genotype-by-environment interactions and can be implemented in any species phenotyped in multiple environments.

@@

  We thank the Brackenridge Field laboratory, the Ladybird Johnson Wildflower Center, and the Juenger laboratory for support with plant care and propagation. This material is based upon work supported in part by the Great Lakes Bioenergy Research Center, U.S. Department of Energy, Office of Science, Office of Biological and Environmental Research under Award Numbers DE-SC0018409 and DE-FC02-07ER64494, the US Department of Energy Awards DE-SC0021126 and DESC0014156 to T.E.J., DE-SC0017883 to D.B.L, NIH grant R35GM151108 and a Pew Scholarship to A.H., National Science Foundation PGRP Awards IOS0922457 and IOS1444533 to T.E.J, and the Long-term Ecological Research Program (DEB 1832042) at the Kellogg Biological Station.

@@

\firstpage[13]{2}

@@

Flowering time, or the transition from vegetative to reproductive growth, is a common subject of GxE research [@blackman2013interacting; @henry2014transitions; @aagren2017adaptive; @brachi2010linkage; @dittmar2014flowering; @li2018genomic; @romero2017study; @bustos-korts2019], a key output of selection driving adaptation to local environments [@unterseer2016; @andres2012genetic; @wadgymar2017identifying; @blumel2015flowering; @fournier-level2016], and a selection target for crop improvement to adapt crops to local or future environments @jung2009flowering. Changing flowering responsiveness to photoperiod cues has allowed geographic range expansion and increased yields in several cereal species [@unterseer2016; @turner2005pseudo; @faure2012mutation; @hung2012zmcct; @zakhrabekova2012induced; @yang2013oself3] and other crops [@pin2012multifaceted; @weller2019parallel]. Recent statistical advances in studying phenological GxE have involved determining critical environmental indices before the phenological event occurs, such as photothermal time within a critical growth window @li2018genomic. However, most studies of flowering GxE focus on finding a single, best fitting form of genotype-environment covariance, despite the key expectation that different genetic subpopulations, and even different genomic regions, have likely evolved distinct patterns of GxE [@Weine2023.06.21.545998]. Additionally, despite theoretical predictions that local adaptation should involve trade-offs caused by antagonistic pleiotropy, or alleles with effects with opposing fitness outcomes [@levene1953genetic; @felsenstein1976theoretical; @kawecki2004conceptual; @hedrick1986genetic], previous empirical work has found limited evidence of trade-offs caused by this form of GxE [@des2013genotype; @wadgymar2017identifying; @millet2016]. However, this work has been limited by a known statistical bias that reduced detection of genetic effects that differ in sign [@des2013genotype; @anderson2013genetic; @anderson2011evolutionary]. Thus, despite substantial interest in the frequencies of various forms of GxE, the frequency of sign-changing GxE relative to other forms of GxE remains unknown.

@@

We expected patterns of GxE and GxWeather to vary at the locus, subpopulation, and trait level. Thus, we were interested in modeling the different ways polygenic effects might covary across our eight common gardens[@weine2024]. We jointly re-estimated the genetic effects of a subset of SNPs across all eight common gardens using *mash* models, which can flexibly capture many modes of covariance across gardens (SI Appendix, Sections S1-S5, Datasets 1-6). To obtain genetic effects, we used subsets of effects from garden- and subpopulation GWAS (SI Appendix, Section S2-S3; Table S1). To obtain covariance structures, we specified both covariance structures introduced in the initial *mash* manuscript (e.g., garden-specific effects; equal effects at all gardens, @fig-covar A), and GxWeather covariance matrices estimated from the covariance of empirical weather patterns at each garden at specific times before the phenological event (Table 1; SI Appendix, Section S1). Then, we used a greedy algorithm to select covariance matrices from each model's set that significantly improved the model likelihood (SI Appendix, Section S4; Table S2). Finally, we fit *mash* models using the selected set of covariance matrices on a subset of relatively unlinked (r^2^ \< 0.2), 'strong' genetic effects with low *p*-values (SI Appendix, Section S5).

@@

For green-up date, SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (@fig-covar B). In total, 65% of the posterior weight of strong SNP effects in the *mash* model of Gulf green-up fell on a covariance matrix constructed using the covariance of day length 14 days prior to the date of green-up. The covariance matrix for this weather cue was visually similar to the observed pattern of phenotypic correlation for the timing of vegetative growth in the Gulf subpopulation (@fig-map B; @fig-covar A). *Mash* models of the timing of vegetative growth in the Midwest subpopulation and Both subpopulations did not include this GxWeather covariance. The Midwest had 0.15% weight on a GxWeather covariance matrix of cumulative GDD for the 28d prior to green-up, a GxWeather matrix that was visually similar to the Midwest's observed pattern of phenotypic correlation (@fig-map B). Both subpopulations had non-zero weights on two additional GxWeather covariance types, average temperature one day prior to green-up, and the day length change in the day prior to green-up (@fig-covar B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations observed in Both subpopulations (@fig-map B; @fig-covar A). The Gulf subpopulation and Both subpopulations had substantially more mass on GxWeather covariance matrices than the Midwest population (@fig-covar C) for the onset of vegetative growth.

For flowering date, distinct GxWeather covariance structures captured covariance in effect sizes for SNPs in the Gulf and Midwest subpopulations. 33% of SNP effects on flowering in the Gulf subpopulation covaried with cumulative rainfall in the seven days prior to flowering (@fig-covar D). 22.6% of SNP effects on flowering in the Midwest subpopulation covaried with day length change in the two days prior to flowering (@fig-covar D), which had negative covariances between Texas and North gardens. Neither of these covariance matrices significantly improved the log-likelihood of the *mash* model of Both subpopulations, and no GxWeather covariance matrices had non-zero mass in models of Both subpopulations (@fig-covar E). In five of the six *mash* models of strong effects, the GxWeather covariance matrices captured a minority of the posterior weights of the strong effects (@fig-covar C,E); in these five models, the majority of this mass was on various canonical covariance matrices. These matrices included simple heterozygosity, with intermediate, positive covariances between all gardens, and single effect matrices with garden-specific effects. Re-estimation of genetic effects with *mash* allowed us to identify and quantify the fraction of loci exhibiting GxWeather patterns; a minority of GxE was GxWeather covariance, and the majority of GxE covariance in these models is driven by other, unknown drivers of GxE.

@@

We expected to observe common rank-changing GxE at the level of individual loci as this is a key theoretical prediction of local adaptation [@felsenstein1976theoretical; @levene1953genetic; @kawecki2004conceptual; @hedrick1986genetic]. Previous empirical work has found limited evidence of trade-offs caused by antagonistic pleiotropy; however, this work had a known statistical bias reducing the detection of effects that differed in sign [@wadgymar2017identifying; @des2013genotype; @millet2016]. To determine the frequency of rank-changing GxE, we used the local false sign rate (lfsr), an analogue of the local false discovery rate that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for a sign change between gardens carry an equal statistical burden to those for effects with the same sign. We separated kinds of effects at the level of individual loci into SNP effects that differ in sign between gardens (effects with rank-changing GxE between gardens), SNP effects that differ in magnitude (effects that are large in one garden or region and smaller in others), and SNP effects that are indistinguishable in two gardens (similarly large or small in comparisons between gardens) (@fig-effects A).

@@

As the climate and the natural environment change, it is increasingly critical to understand how patterns of plant-environment interactions will change in response. To do this, we must understand the current patterns of trait covariation across environments, the genetic underpinnings of these patterns, and the cases where this covariation can be altered through selection at individual loci. Here, we demonstrate that we can associate multiple patterns of GxWeather with specific genomic regions using a switchgrass diversity panel grown at eight common gardens. We assigned genetic effects to both GxWeather patterns with interpretable weather-based cues, and to unmeasured, site-based patterns. We used this approach to study GxWeather for the timings of vegetative and reproductive development in the deeply genetically diverged Gulf and Midwest subpopulations of switchgrass.

Our analysis of the timing of vegetative and reproductive growth revealed many alleles with rank-changing GxE, or sign changes in their effects between the Texas and North gardens (@fig-effects B). The Gulf & Both subpopulations had rank-changing GxE for green-up date, while the Midwest subpopulation had rank-changing GxE for flowering date. As phenological timings are major components of plant fitness, this result supports theoretical models that local adaptation should involve trade-offs due to antagonistic pleiotropy at the level of individual loci [@kawecki2004conceptual; @felsenstein1976theoretical; @levene1953genetic; @hedrick1986genetic]. Experimental designs in local adaptation research now often use more than two field sites and a wide range of genetic variation; *mash* and the local false sign rate facilitate the analysis of these experimental designs. We use *mash* to determine that rank-changing GxE for phenological traits is common in small genomic regions [@wadgymar2017identifying; @savolainen2013ecological; @lowry2019qtl].

Our analysis of the timing of flowering showed that the Gulf and Midwest subpopulations have distinct GxWeather: flowering timing in the Midwest subpopulation has photoperiod-related genetic variation, in that flowering timing covaries with a day length change signal two days before flowering occurs. In contrast, the Gulf subpopulation does not have genetic variation in flowering that covaries with a photoperiod cue. Instead, the Gulf subpopulation has genetic variation in flowering that covaries with the rainfall that occurs in the week prior to flowering. Three genomic regions affecting flowering that we re-estimated across all eight sites were also supported by QTL from an independent mapping population at these sites (@fig-qtl B). Models combining both subpopulations showed less signal, perhaps due to the distinctness of cues in both subpopulations or to confounding caused by population structure.
